# Inner workings of the framework

```{r}
#| include: false
source("_common.R")
```

Underneath the hood, the `arx_forecaster()` (and all our canned
forecasters) creates and returns an `epi_workflow`. 
Essentially, this is a big S3 object that wraps up the 4 modular steps 
(preprocessing - postprocessing) described in the last chapter.

1. Preprocessor: make transformations to the data before model training
2. Trainer: train a model on data, resulting in a fitted model object
3. Predictor: make predictions, using a fitted model object and processed test data
4. Postprocessor: manipulate or transform the predictions before returning

Let's investigate how these interact with `{tidymodels}` and why it's important
to think of forecasting this way. To have something to play with, we'll continue
to examine the data and an estimated canned forecaster.


```{r demo-workflow}
jhu <- case_death_rate_subset %>% 
  filter(time_value >= max(time_value) - 30)

out_gb <- arx_forecaster(jhu, "death_rate", c("case_rate", "death_rate"),
  boost_tree(mode = "regression", trees = 20))
```

## Preprocessing

Preprocessing is accomplished through a `recipe` (imagine baking a cake) as 
provided in the [`{recipes}`](https://recipes.tidymodels.org) package. 
We've made a few modifications (to handle
panel data) as well as added some additional options. The recipe gives a
specification of how to handle training data. Think of it like a fancified
`formula` that you would pass to `lm()`: `y ~ x1 + log(x2)`. In general, 
there are 2 extensions to the `formula` that `{recipes}` handles: 

  1. Doing transformations of both training and test data that can always be 
  applied. These are things like taking the log of a variable, leading or 
  lagging, filtering out rows, handling dummy variables, etc.
  2. Using statistics from the training data to eventually process test data. 
    This is a major benefit of `{recipes}`. It prevents what the tidy team calls
    "data leakage". A simple example is centering a predictor by its mean. We
    need to store the mean of the predictor from the training data and use that
    value on the test data rather than accidentally calculating the mean of
    the test predictor for centering.
    
A recipe is processed in 2 steps, first it is "prepped". This calculates and
stores any intermediate statistics necessary for use on the test data. 
Then it is "baked"
resulting in training data ready for passing into a statistical model (like `lm`).

We have introduced an `epi_recipe`. It's just a `recipe` that knows how to handle
the `time_value`, `geo_value`, and any additional keys so that these are available
when necessary.

The `epi_recipe` from `out_gb` can be extracted from the result:
```{r}
#| results: false
library(workflows)
library(recipes)
extract_recipe(out_gb$epi_workflow)
```

```{r}
#| message: true
#| echo: false
#| collapse: true
extract_recipe(out_gb$epi_workflow)
```


The "Inputs" are the original `epi_df` and the "roles" that these are assigned.
None of these are predictors or outcomes. Those will be created 
by the recipe when it is prepped. The "Operations" are the sequence of 
instructions to create the cake (baked training data).
Here we create lagged predictors, lead the outcome, and then remove `NA`s.
Some models like `lm` internally handle `NA`s, but not everything does, so we
deal with them explicitly. The code to do this (inside the forecaster) is

```{r}
er <- epi_recipe(jhu) %>%
  step_epi_lag(case_rate, death_rate, lag = c(0, 7, 14)) %>%
  step_epi_ahead(death_rate, ahead = 7) %>%
  step_epi_naomit()
```

While `{recipes}` provides a function `step_lag()`, it assumes that the data
have no breaks in the sequence of `time_values`. This is a bit dangerous, so
we avoid that behaviour. Our `lag/ahead` functions also appropriately adjust the
amount of data to avoid accidentally dropping recent predictors from the test
data.

## The model specification

Users familiar with the `{parsnip}` package will have no trouble here.
Basically, `{parsnip}` unifies the function signature across statistical models.
For example, `lm()` "likes" to work with formulas, but `glmnet::glmnet()` uses
`x` and `y` for predictors and response. `{parsnip}` is agnostic. Both of these
do "linear regression". Above we switched from `lm()` to `xgboost()` without 
any issue despite the fact that these functions couldn't be more different.

```{r, eval = FALSE}
lm(
  formula, data, subset, weights, na.action, method = "qr", 
  model = TRUE, x = FALSE, y = FALSE, qr = TRUE, singular.ok = TRUE, 
  contrasts = NULL, offset, ...)

xgboost(
  data = NULL, label = NULL, missing = NA, weight = NULL, 
  params = list(), nrounds, verbose = 1, print_every_n = 1L, 
  early_stopping_rounds = NULL, maximize = NULL, save_period = NULL, 
  save_name = "xgboost.model", xgb_model = NULL, callbacks = list(), 
  ...) 
```

`{epipredict}` provides a few engines/modules like `flatline()` and 
`quantile_reg()` to power the `flatline_forecaster()` and provide quantile 
regression, but you should be able to use almost any available models
listed [here](https://www.tidymodels.org/find/parsnip/).


To estimate (fit) a preprocessed model, one calls `fit()` on the `epi_workflow`.

```{r}
ewf <- epi_workflow(er, linear_reg()) %>% fit(jhu)
```

## Predicting and Postprocessing (bound together)

To stretch the metaphor of preparing a cake to its natural limits, we have
created postprocessing functionality called "frosting". Much like the recipe,
each postprocessing operation is a "layer" and we "slather" these onto our 
baked cake. To fix ideas, below is the postprocessing `frosting` for 
`arx_forecaster()`

```{r}
#| results: false
extract_frosting(out_gb$epi_workflow)
```

```{r}
#| echo: false
#| results: asis
#| message: true
#| collapse: true
extract_frosting(out_gb$epi_workflow)
```


Here we have 5 layers of frosting. The first generates the forecasts from the test data.
The second uses quantiles of the residuals to create distributional
forecasts. The next two add columns for the date the forecast was made and the
date for which it is intended to occur. Because we are predicting rates, they 
should be non-negative, so the last layer thresholds both predicted values and
intervals at 0. The code to do this (inside the forecaster) is

```{r}
f <- frosting() %>%
  layer_predict() %>%
  layer_residual_quantiles(
    probs = c(.01, .025, seq(.05, .95, by = .05), .975, .99), 
    symmetrize = TRUE) %>%
  layer_add_forecast_date() %>%
  layer_add_target_date() %>%
  layer_threshold(starts_with(".pred"))
```

At predict time, we add this object onto the `epi_workflow` and call `predict()`

```{r, warning=FALSE}
test_data <- get_test_data(er, jhu)
ewf %>% add_frosting(f) %>% predict(test_data)
```

The above `get_test_data()` function examines the recipe and ensures that enough
test data is available to create the necessary lags and produce a prediction
for the desired future time point (after the end of the training data). This mimics
what would happen if `jhu` contained the most recent available historical data and
we wanted to actually predict the future. We could have instead used any test data
that contained the necessary predictors.

:::{.callout-note}
In the predictions above, you'll see a `time_value` column. That's because we 
could use **any training data**. We happened to use training data corresponding
to the most recent available, and it's lags. But we could have instead used
last week's or we could use the data that arrives next year, or we could use multiple
`time_values` for multiple locations. This is completely allowed, though not
necessarily what you expect.

In production forecasting, you'd probably reestimate the model and produce new
predictions whenever new data arrives. This is exactly what all the canned 
forecasters we provide do. So those strip out the `time_value` column.

But the next most likely procedure would be
to feed your previously estimated model (without refitting) the new data.
To do this, you'd just call `get_test_data()` on that new data. And the 
`time_value` would still be the same as your `forecast_date`.

Getting many forecasts (multiple `time_values`) for each location, is not
exactly a typical desire in this context. But it's also not unheard of, so
it is possible (and analogous to standard, non-time series forecasting). 
:::


## Conclusion

Internally, we provide some canned forecaster functions to create reasonable forecasts. 
But ideally, a user could create their own forecasters by building up the 
components we provide. In other chapters, we try to walk through some of these
customizations. 

To illustrate everything above, here is (roughly) the code for the 
`arx_forecaster()` to predict the death rate, 1 week ahead:

```{r}
r <- epi_recipe(jhu) %>%
  step_epi_ahead(death_rate, ahead = 7) %>%
  step_epi_lag(case_rate, death_rate, lag = c(0, 7, 14)) %>%
  step_epi_naomit()

latest <- get_test_data(r, jhu) 

f <- frosting() %>%
  layer_predict() %>%
  layer_residual_quantiles() %>%
  layer_add_forecast_date() %>%
  layer_add_target_date() %>%
  layer_threshold(starts_with(".pred"))

eng <- linear_reg()
wf <- epi_workflow(r, eng, f) %>% fit(jhu)
preds <- predict(wf, latest)
```
The code for `arx_forecaster()` simply generalizes this, passing along arguments as needed.

```{r}
preds
```


