# Examples of Preprocessing and Models

```{r}
#| echo: false
source("_common.R")
```


## Introduction 

The `{epipredict}` package uses the `{tidymodels}` framework, namely 
`{recipes}` for `{dplyr}`-like pipeable sequences of feature engineering and `{parsnip}` for a unified interface to a range of models. 

`epipredict` has additional customized feature engineering and preprocessing 
steps that specifically work with panel data in this context, for example,
`step_epi_lag()`, `step_population_scaling()`, 
`step_epi_naomit()`. They can be used along with most
steps from the `recipes` package for more feature engineering. 

In this vignette, we will illustrate some examples of how to use `epipredict`
with `recipes` and `parsnip` for different purposes of 
epidemiological forecasting.
We will focus on basic autoregressive models, in which COVID cases and 
deaths in the near future are predicted using a linear combination of cases
and deaths in the near past.

The remaining vignette will be split into three sections. In the first
section, we 
will use a Poisson regression to predict death counts. In the second section,
we will use a linear regression to predict death rates. Last but not least, we
will create a classification model for hotspot predictions. 

```{r, warning=FALSE, message=FALSE}
library(epidatr)
library(epipredict)
library(recipes)
library(workflows)
library(poissonreg)
library(epidatasets)
```

## Poisson Regression 

During COVID-19, the U.S. Centers for Disease Control and Prevention (CDC) 
collected models
and forecasts to characterize the state of an outbreak and its course. They use
it to inform public health decision makers on potential consequences of 
deploying control measures.

One of the outcomes that the CDC forecasts is [death counts from COVID-19](https://www.cdc.gov/coronavirus/2019-ncov/science/forecasting/forecasting-us.html).
Although there are many state-of-the-art models, we choose to use Poisson 
regression, the textbook example for modeling count data, as an illustration
for using the `epipredict` package with other existing `tidymodels` packages. 

The (folded) code below gives the necessary commands to download this data
`counts_subset` from the Delphi Epidata API, but it is also built into the `{epidatasets}` package.

```{r poisson-reg-data}
#| eval: false
#| code-fold: true
geos <- c("ca", "fl", "tx", "ny", "nj")
x <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = geos) %>%
  fetch() %>%
  select(geo_value, time_value, cases = value)

y <- covidcast(
  data_source = "jhu-csse",
  signals = "deaths_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = geos) %>%
  fetch() %>%
  select(geo_value, time_value, deaths = value)

counts_subset <- full_join(x, y, by = c("geo_value", "time_value")) %>%
  as_epi_df()
```

```{r}
data(counts_subset, package = "epidatasets")
```

The `epidatasets::counts_subset` dataset
contains the number of confirmed cases and deaths from June 4, 2021 to 
Dec 31, 2021 in some U.S. states. 

We wish to predict the 7-day ahead death counts with lagged cases and deaths.
Furthermore, we will include the categorical variable, *states*, as dummy variables. Using different 
intercept coefficients, we can allow for an intercept shift between states.

One possible model takes the form
\begin{aligned}
\log\left( \mu_{t+7} \right) &{}= \beta_0 + \delta_1 s_{\text{state}_1} +
\delta_2 s_{\text{state}_2} + \cdots +  \nonumber \\ &\quad\beta_1 \text{deaths}_{t} + 
\beta_2 \text{deaths}_{t-7}  + \beta_3 \text{cases}_{t} + 
\beta_4 \text{cases}_{t-7},
\end{aligned}
where $\mu_{t+7} = \mathbb{E}(\text{deaths}_{t+7})$, and $\text{deaths}_{t+7}$
is assumed to follow a Poisson distribution with mean $\mu_{t+7}$;
$s_{\text{state}}$ are dummy variables for each state and take values of either
0 or 1.

Preprocessing steps will be performed to prepare the
data for model fitting. But before diving into them, it will be helpful to understand what `roles` are in the `recipes` framework. 

---

#### Aside on `recipes` {.unnumbered}

`{recipes}` can assign one or more roles to each column in the data. The roles 
are not restricted to a predefined set; they can be anything. 
For most conventional situations, they are typically “predictor” and/or 
"outcome". Additional roles enable targeted `step_*()` operations on specific 
variables or groups of variables.

In our case, the role `predictor` is given to explanatory variables on the
right-hand side of the model (in the equation above). 
The role `outcome` is the response variable 
that we wish to predict. `geo_value` and `time_value` are predefined roles 
that are unique to the `epipredict` package. Since we work with `epi_df` 
objects, all datasets should have `geo_value` and `time_value` passed through
automatically with these two roles assigned to the appropriate columns in the data.
 
The `recipes` package also allows [manual alterations of roles](https://recipes.tidymodels.org/reference/roles.html) 
in bulk. There are a few handy functions that can be used together to help us 
manipulate variable roles easily. 

> `update_role()` alters an existing role in the recipe or assigns an initial role 
> to variables that do not yet have a declared role.
> 
> `add_role()` adds an additional role to variables that already have a role in 
> the recipe, without overwriting old roles.
> 
> `remove_role()` eliminates a single existing role in the recipe.

#### End aside {.unnumbered}

---

Notice in the following preprocessing steps, we used `add_role()` on 
`geo_value_factor` since, currently, the default role for it is `raw`, but
we would like to reuse this variable as a `predictor`.

```{r}
counts_subset <- counts_subset %>%
  mutate(geo_value_factor = as.factor(geo_value)) %>%
  as_epi_df()

epi_recipe(counts_subset)

r <- epi_recipe(counts_subset) %>%
  add_role(geo_value_factor, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  ## Occasionally, data reporting errors / corrections result in negative
  ## cases / deaths
  step_mutate(cases = pmax(cases, 0), deaths = pmax(deaths, 0)) %>%  
  step_epi_lag(cases, deaths, lag = c(0, 7)) %>%
  step_epi_ahead(deaths, ahead = 7, role = "outcome") %>%
  step_epi_naomit()
```

After specifying the preprocessing steps, we will use the `{parsnip}` package for
modeling and producing the prediction for death count, 7 days after the
latest available date in the dataset. 

```{r}
latest <- get_test_data(r, counts_subset)

wf <- epi_workflow(r, parsnip::poisson_reg()) %>%
  fit(counts_subset)

predict(wf, latest) %>% filter(!is.na(.pred))
```

Note that the `time_value` corresponds to the date(s) in the 
test set `latest`, **NOT** to the target date of the forecast (`r max(latest$time_value) + 7`). Had we used different data for predictions,
we would have gotten different `time_value`'s.

Let's take a look at the fit:
```{r}
extract_fit_engine(wf)
```

Alternative forms of Poisson regression or particular computational approaches
can be applied via arguments to `poisson_reg()` for some common
settings, and by using `set_engine()` to use a specific Poisson
regression engine and to provide additional engine-specific customization.



## Linear Regression 

For COVID-19, the CDC required submission of case and death count predictions. 
However, the Delphi Group preferred to train on rate data instead, because it 
puts different locations on a similar scale (eliminating the need for location-specific intercepts). 
We can use a linear regression to predict the death rates and use state
population data to scale the rates to counts.[^pois] We will do so using
`layer_population_scaling()` from the `epipredict` package. (We could also use
`step_population_scaling()` to prepare rate data
from count data in the preprocessing recipe.)

[^pois]: We could continue with the Poisson model, but we'll switch to the Gaussian likelihood just for simplicity.

Additionally, when forecasts are submitted, prediction intervals should be 
provided along with the point estimates. This can be obtained via postprocessing
using
`layer_residual_quantiles()`. It is worth pointing out, however, that 
`layer_residual_quantiles()` should be used before population scaling or else 
the transformation will make the results uninterpretable. 

We wish, now, to predict the 7-day ahead death counts with lagged case rates and death
rates, along with some extra behaviourial predictors. Namely, we will use survey data
from [COVID-19 Trends and Impact Survey](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html#behavior-indicators).

The survey data provides the estimated percentage of people who wore a mask for 
most or all of the time while in public in the past 7 days and the estimated 
percentage of respondents who reported that all or most people they encountered 
in public in the past 7 days maintained a distance of at least 6 feet. 

```{r}
#| eval: false
#| code-fold: true
# Download the raw data as used in {epidatasets}
behav_ind_mask <- covidcast(
  data_source = "fb-survey",
  signals = "smoothed_wwearing_mask_7d",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = geos)  %>%
  fetch() %>%
  select(geo_value, time_value, masking = value)

behav_ind_distancing <- covidcast(
  data_source = "fb-survey",
  signals = "smoothed_wothers_distanced_public",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = geos)  %>%
  fetch() %>%
  select(geo_value, time_value, distancing = value) 

ctis_covid_behaviours <- behav_ind_mask %>%
  full_join(behav_ind_distancing, by = c("geo_value", "time_value")) 
```

```{r}
data(ctis_covid_behaviours, package = "epidatasets")
pop_dat <- state_census %>% select(abbr, pop)
```

State-wise population data from the 2019 U.S. Census is
available from `epipredict` and will be used in `layer_population_scaling()`. 



Rather than using raw mask-wearing / social-distancing metrics, for the sake
of illustration, we'll convert both into categorical predictors.

```{r, echo=FALSE, message=FALSE,fig.align='center', fig.width=6, fig.height=4}
#| code-fold: true
ctis_covid_behaviours %>%
  pivot_longer(masking:distancing) %>%
  ggplot(aes(value, fill = geo_value)) +
  geom_density(alpha = 0.5) +
  scale_fill_brewer(palette = "Set1", name = "") +
  theme_bw() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(c(0, .05))) +
  facet_wrap(~ name, scales = "free") +
  theme(legend.position = "bottom")
```

We will take a subset of death rate and case rate data from the built-in dataset 
`case_death_rate_subset`.

```{r}
jhu <- filter(
  case_death_rate_subset,
  time_value >= "2021-06-04", 
  time_value <= "2021-12-31",
  geo_value %in% c("ca","fl","tx","ny","nj")
)
```

Preprocessing steps will again rely on functions from the `epipredict` package 
as well as the `recipes` package.
There are also many functions in the `recipes` package that allow for 
[scalar transformations](https://recipes.tidymodels.org/reference/#step-functions-individual-transformations),
such as log transformations and data centering. In our case, we will 
center the numerical predictors to allow for a more meaningful interpretation of
the intercept. 

```{r}
jhu <- jhu %>%
  mutate(geo_value_factor = as.factor(geo_value)) %>%
  left_join(ctis_covid_behaviours, by = c("geo_value", "time_value")) %>%
  as_epi_df()
            
r <- epi_recipe(jhu) %>%
  add_role(geo_value_factor, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  step_epi_lag(case_rate, death_rate, lag = c(0, 7, 14)) %>%
  step_mutate(
    masking = cut_number(masking, 5), 
    distancing = cut_number(distancing, 5)
  ) %>%
  step_epi_ahead(death_rate, ahead = 7, role = "outcome") %>%
  step_center(contains("lag"), role = "predictor") %>%
  step_epi_naomit()
```

As a sanity check we can examine the structure of the training data:
```{r, warning = FALSE}
glimpse(bake(prep(r, jhu), jhu))
```

Before directly predicting the results, we need to add postprocessing layers to
obtain the death counts instead of death rates. Note that the rates used so
far are "per 100K people" rather than "per person". We'll also use quantile
regression with the `quantile_reg` engine rather than ordinary least squares
to create median predictions and a 90% prediction interval.

```{r, warning=FALSE}
f <- frosting() %>%
  layer_predict() %>%
  layer_add_target_date("2022-01-07") %>% 
  layer_add_forecast_date() %>%
  layer_threshold(.pred, lower = 0) %>%
  layer_quantile_distn() %>%
  layer_point_from_distn() %>%
  layer_naomit(.pred) %>%
  layer_population_scaling(
    contains(".pred"),
    df = pop_dat, 
    rate_rescaling = 1e5,
    by = c("geo_value" = "abbr"), 
    df_pop_col = "pop")

wf <- epi_workflow(r, quantile_reg(tau = c(.05, .5, .95))) %>%
  fit(jhu) %>%
  add_frosting(f)

latest <- get_test_data(recipe = r, x = jhu)
p <- predict(wf, latest) %>% select(-time_value) %>% as_tibble()
p
```

The columns marked `*_scaled` (unfortunately, some of these
are hidden above) 
have been rescaled to the correct units, in this
case `deaths` rather than deaths per 100K people (these remain in `.pred`).

To look at the prediction intervals:
```{r}
p %>%
  select(geo_value, target_date, .pred_scaled, .pred_distn_scaled) %>%
  pivot_quantiles(.pred_distn_scaled)
```


Last but not least, let's take a look at the regression fit and check the 
coefficients:
```{r, echo =FALSE}
extract_fit_engine(wf)
```

## Classification

Sometimes it is preferable to create a predictive model for surges or upswings
rather than for raw values. In this case,
the target is to predict if the future will have increased case rates (denoted `up`),
decreased case rates (`down`), or flat case rates (`flat`) relative to the current
level. Such models may be 
referred to as "hotspot prediction models". We will follow the analysis 
in [McDonald, Bien, Green, Hu, et al.](#references) but extend the application
to predict three categories instead of two. 

Hotspot prediction uses a categorical outcome variable defined in terms of the 
relative change of $Y_{\ell, t+a}$ compared to $Y_{\ell, t}$. 
Where $Y_{\ell, t}$ denotes the case rates in location $\ell$ at time $t$. 
We define the response variables as follows:

$$
 Z_{\ell, t}=
    \begin{cases}
      \text{up}, & \text{if}\ Y^{\Delta}_{\ell, t} > 0.25 \\ 
      \text{down}, & \text{if}\  Y^{\Delta}_{\ell, t} < -0.20\\
      \text{flat}, & \text{otherwise}
    \end{cases}
$$

where $Y^{\Delta}_{\ell, t} = (Y_{\ell, t}- Y_{\ell, t-7})\ /\ (Y_{\ell, t-7})$. 
We say location $\ell$ is a hotspot at time $t$ when $Z_{\ell,t}$ is 
`up`, meaning the number of newly reported cases over the past 7 days has 
increased by at least 25% compared to the preceding week. When $Z_{\ell,t}$ 
is categorized as `down`, it suggests that there has been at least a 20% 
decrease in newly reported cases over the past 7 days (a 20% decrease is the inverse of a 25% increase). Otherwise, we will 
consider the trend to be `flat`. 

The expression of the multinomial regression we will use is as follows:
$$
\pi_{j}(x) = \text{Pr}(Z_{\ell,t} = j|x) = \frac{e^{g_j(x)}}{1 + \sum_{k=0}^2 g_j(x) }
$$
where $j$ is either down, flat, or up

$$
\begin{aligned}
g_{\text{down}}(x) &= 0,\\
g_{\text{flat}}(x) &= 
\log\left(\frac{Pr(Z_{\ell,t}=\text{flat}|x)}{Pr(Z_{\ell,t}=\text{down}|x)}\right) = 
\beta_{10} + \beta_{11}t + \delta_{10} s_{\text{state}_1} +
\delta_{11} s_{\text{state}_2} + \cdots \nonumber \\
&\quad +\ \beta_{12} Y^{\Delta}_{\ell, t} +
\beta_{13} Y^{\Delta}_{\ell, t-7}, \\
g_{\text{flat}}(x) &= \log\left(\frac{Pr(Z_{\ell,t}=\text{up}|x)}{Pr(Z_{\ell,t}=\text{down}|x)}\right) = 
\beta_{20} + \beta_{21}t + \delta_{20} s_{\text{state}_1} +
\delta_{21} s_{\text{state}_2} + \cdots \nonumber \\
&\quad +\ \beta_{22} Y^{\Delta}_{\ell, t} +
\beta_{23} Y^{\Delta}_{\ell, t-7}.
\end{aligned}
$$


Preprocessing steps are similar to the previous models with an additional step 
of categorizing the response variables. Again, we will use a subset of death rate and case rate data from our built-in dataset 
`epipredict::case_death_rate_subset`.
```{r}
jhu_rates <- case_death_rate_subset %>%
  dplyr::filter(
    time_value >= "2021-06-04", 
    time_value <= "2021-12-31",
    geo_value %in% c("ca","fl","tx","ny","nj")) %>%
  mutate(geo_value_factor = as.factor(geo_value)) 

r <- epi_recipe(jhu_rates) %>%
  add_role(time_value, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  step_growth_rate(case_rate, role = "none", prefix = "gr_") %>%
  step_epi_lag(starts_with("gr_"), lag = c(0, 7, 14)) %>%
  step_epi_ahead(starts_with("gr_"), ahead = 7, role = "none") %>%
  # note recipes::step_cut() has a bug in it, or we could use that here
  step_mutate(
    response = cut(
      ahead_7_gr_7_rel_change_case_rate, 
      breaks = c(-Inf, -0.2, 0.25, Inf) / 7, # division gives weekly not daily
      labels = c("down", "flat", "up")),
    role = "outcome"
  ) %>%
  step_rm(has_role("none"), has_role("raw")) %>%
  step_epi_naomit()
```

We will fit the multinomial regression and examine the predictions:

```{r, warning=FALSE}
wf <- epi_workflow(r, parsnip::multinom_reg()) %>%
  fit(jhu_rates)

latest <- get_test_data(recipe = r, x = jhu_rates)
predict(wf, latest) %>% filter(!is.na(.pred_class))
```

We can also look at the estimated coefficients and model summary information:
```{r}
extract_fit_engine(wf)
```

One could also use a formula in `epi_recipe()` to achieve the same results as 
above. However, only one of `add_formula()`, `add_recipe()`, or 
`workflow_variables()` can be specified. For the purpose of demonstrating 
`add_formula` rather than `add_recipe`, we will `prep` and `bake` our recipe to
return a `data.frame` that could be used for model fitting.
```{r}
b <- bake(prep(r, jhu_rates), jhu_rates)

epi_workflow() %>%
  add_formula(
    response ~ geo_value + time_value + lag_0_gr_7_rel_change_case_rate +
      lag_7_gr_7_rel_change_case_rate + lag_14_gr_7_rel_change_case_rate
  ) %>%
  add_model(parsnip::multinom_reg()) %>%
  fit(data = b)
```

<!--

## Benefits of Lagging and Leading in `epipredict`

The `step_epi_ahead` and `step_epi_lag` functions in the `epipredict` package
is handy for creating correct lags and leads for future predictions. 

Let's examine what happens with one month of data for one location:
```{r}
ex <- filter(
  case_death_rate_subset, 
  time_value >= "2021-12-01", 
  time_value <= "2021-12-31",
  geo_value == "ca"
)
ex
```

We want to predict death rates on `r max(ex$time_value) + 7`, which is 7 days 
ahead of the latest available date in our dataset. 

We will compare two methods of trying to create lags and leads:
```{r}
prepped_epi <- epi_recipe(ex) %>%
  step_epi_lag(case_rate, death_rate, lag = c(0, 7, 14)) %>%
  step_epi_ahead(death_rate, ahead = 7) %>%
  step_epi_naomit() %>%
  prep()

baked_epi <- bake(prepped_epi, ex)
baked_epi


prepped_rec <- epi_recipe(ex) %>%
  step_lag(case_rate, death_rate, lag = c(7, 14)) %>% # lags must be positive
  step_mutate(lag_0_death_rate = death_rate, lag_0_case_rate = case_rate) %>%
  step_mutate(ahead_7_death_rate = lead(death_rate, 7), role = "outcome") %>%
  step_naomit(all_predictors(), all_outcomes()) %>%
  prep()

baked_rec <- bake(prepped_rec, ex)
baked_rec
```

Notice the difference in number of rows `b1` and `b2` returns. This is because 
the second version, the one that doesn't use `step_epi_ahead` and `step_epi_lag`,
has omitted dates compared to the one that used the `epipredict` functions.
```{r}
dates_used_by_epi <- baked_epi %>% 
  select(-ahead_7_death_rate) %>% 
  na.omit() %>% 
  pull(time_value)
dates_used_by_epi

dates_used_by_rec <- baked_rec %>% 
  select(-ahead_7_death_rate) %>% 
  na.omit() %>% 
  pull(time_value)
dates_used_by_rec
```

The model that is trained based on the `{recipes}` functions will predict 7 days 
ahead from `r max(dates_used_by_rec)`
instead of 7 days ahead from `r max(dates_used_by_epi)`.

-->
