# Sliding computations {#sec-sliding}

A central tool in the `{epiprocess}` package is `epi_slide()`, which is based on
the powerful functionality provided in the
[`slider`](https://cran.r-project.org/web/packages/slider) package. In
`epiprocess`, to "slide" means to apply a computation---represented as a
function or formula---over a sliding/rolling data window. The function always
applies the slide inside each group and the grouping is assumed to be across all
group keys of the `epi_df` (this is the grouping used by default if you do not
group the `epi_df` with a `group_by()`).

By default, the `.window_size` units depend on the `time_type` of the `epi_df`,
which is determined from the types in the `time_value` column of the `epi_df`.
See the "Details" in `epi_slide()` for more.

As in getting started guide, we'll fetch daily reported COVID-19 cases from CA,
FL, NY, and TX (note: here we're using new, not cumulative cases) using the
[`epidatr`](https://github.com/cmu-delphi/epidatr) package, and then convert
this to `epi_df` format.

```{r}
#| include: false
source("_common.R")
```

```{r, message = FALSE, warning=FALSE}
library(epidatr)
library(epiprocess)
library(epipredict)
```

The example data we'll use is part of the package and has 2,684 rows and 3 columns.

```{r}
data(jhu_csse_daily_subset)
edf <- jhu_csse_daily_subset %>%
  select(geo_value, time_value, cases) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

## Optimized rolling mean and sums

For the two most common sliding operations, we offer two optimized versions:
`epi_slide_mean()` and `epi_slide_sum()`. This example gets the 7-day trailing
average of the daily cases. Note that the name of the column(s) that we want to
average is specified as the first argument of `epi_slide_mean()`.

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide_mean("cases", .window_size = 7, na.rm = TRUE) %>%
  ungroup() %>%
  head(10)
```

Note that we passed `na.rm = TRUE` to `data.table::frollmean()` via `...` to
`epi_slide_mean`.

The following computes the 7-day trailing sum of daily cases (and passed `na.rm`
to `data.table::frollsum()` similarly):

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide_sum("cases", .window_size = 7, na.rm = TRUE) %>%
  ungroup() %>%
  head(10)
```

## General sliding with a formula

The previous computations can also be performed using `epi_slide()`, which can
be used for more general sliding computations (but is much slower for the
specific cases of mean and sum).

The same 7-day trailing average of daily cases can be computed by passing in a
formula for the first argument of `epi_slide()`:

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide(~ mean(.x$cases, na.rm = TRUE), .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

If your formula returns a data.frame, then the columns of the data.frame
will be unpacked into the resulting `epi_df`. For example, the following
computes the 7-day trailing average of daily cases and the 7-day trailing sum of
daily cases:

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide(
    ~ data.frame(cases_mean = mean(.x$cases, na.rm = TRUE), cases_sum = sum(.x$cases, na.rm = TRUE)),
    .window_size = 7
  ) %>%
  ungroup() %>%
  head(10)
```

Note that this formula has access to all non-grouping columns present in the
original `epi_df` object and must refer to them with the prefix `.x$...`. As we
can see, the function `epi_slide()` returns an `epi_df` object with a new column
appended that contains the results (from sliding), named `slide_value` as the
default.

Some other information is available in additional variables:

* `.group_key` is a one-row tibble containing the values of the grouping
  variables for the associated group
* `.ref_time_value` is the reference time value the time window was based on

```{r}
# Returning geo_value in the formula
edf %>%
  group_by(geo_value) %>%
  epi_slide(~ .x$geo_value[[1]], .window_size = 7) %>%
  ungroup() %>%
  head(10)

# Returning time_value in the formula
edf %>%
  group_by(geo_value) %>%
  epi_slide(~ .x$time_value[[1]], .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

While the computations above do not look very useful, these can be used as
building blocks for computations that do something different depending on the
geo_value or ref_time_value.

## Slide the tidy way

Perhaps the most convenient way to setup a computation in `epi_slide()` is to
pass in an expression for tidy evaluation. In this case, we can simply define
the name of the new column directly as part of the expression, setting it equal
to a computation in which we can access any columns of `.x` by name, just as we
would in a call to `dplyr::mutate()`, or any of the `dplyr` verbs. For example:

```{r}
slide_output <- edf %>%
  group_by(geo_value) %>%
  epi_slide(cases_7dav = mean(cases, na.rm = TRUE), .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

In addition to referring to individual columns by name, you can refer to
`epi_df` time window as `.x` (`.group_key` and `.ref_time_value` are still
available). Also, the tidyverse "pronouns" `.data` and `.env` can also be used
if you need distinguish between the data and environment.

As a simple sanity check, we visualize the 7-day trailing averages computed on
top of the original counts:

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
library(ggplot2)
theme_set(theme_bw())

ggplot(slide_output, aes(x = time_value)) +
  geom_col(aes(y = cases, fill = geo_value), alpha = 0.5, show.legend = FALSE) +
  geom_line(aes(y = cases_7dav, col = geo_value), show.legend = FALSE) +
  facet_wrap(~geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

As we can see from the top right panel, it looks like Texas moved to weekly
reporting of COVID-19 cases in summer of 2021.

## Slide with a function

We can also pass a function to the second argument in `epi_slide()`. In this
case, the passed function `.f` must have the form `function(x, g, t, ...)`,
where

- "x" is an epi_df with the same column names as the archive's `DT`, minus
  the `version` column
- "g" is a one-row tibble containing the values of the grouping variables
for the associated group
- "t" is the ref_time_value for the current window
- "..." are additional arguments

Recreating the last example of a 7-day trailing average:

```{r}
x <- edf %>%
  group_by(geo_value) %>%
  epi_slide(function(x, g, t) mean(x$cases, na.rm = TRUE), .window_size = 7, .new_col_name = "cases_7dav") %>%
  ungroup()
x %>%
  head(10)
```

```{r, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6}
#| code-fold: true
cols <- RColorBrewer::brewer.pal(7, "Set1")[-6]
ggplot(x, aes(x = time_value)) +
  geom_col(aes(y = cases, fill = geo_value), alpha = 0.5,
           show.legend = FALSE) +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  geom_line(aes(y = cases_7dav, col = geo_value), show.legend = FALSE) +
  scale_fill_manual(values = cols) +
  scale_color_manual(values = cols) +
  facet_wrap(~ geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %Y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

As we can see from the center top panel, it looks like Florida moved to weekly
reporting of COVID-19 cases in summer of 2021, while California occasionally
reported negative cases counts!

## Running a local forecaster {#sec-local-forecaster}

As a more complex example, we preview some of the functionality of
`{epipredict}` described in future chapters, and use a forecaster based on a
local (in time) autoregression or "AR model". AR models can be fit in numerous
ways (using base R functions and various packages), but here we the
`arx_forecaster()`, implemented in `{epipredict}` both provides a more advanced
example of sliding a function over an `epi_df` object, and it allows us to be a
bit more flexible in defining a *probabilistic* forecaster: one that outputs not
just a point prediction, but a notion of uncertainty around this. In particular,
our forecaster will output a point prediction along with an 90\% uncertainty
band, represented by a predictive quantiles at the 5\% and 95\% levels (lower
and upper endpoints of the uncertainty band).

The function signature below, is a probabilistic AR forecaster. The
`lags` argument indicates which lags to use in the model, and `ahead` indicates
how far ahead in the future to make forecasts (both are encoded in terms of the
units of the `time_value` column; so, days, in the working `epi_df` being
considered in this vignette).

```{r eval=FALSE}
arx_forecaster <- function(
    epi_df,
    outcome, # the outcome column name in `epi_df`
    predictors, # a character vector, containing 1 or more predictors in `epi_df`
    trainer = quantile_reg(),
    args_list = arx_args_list(
      lags = c(0, 7, 14),
      ahead = 7,
      quantile_levels = c(0.05, 0.95)
    )
) {
  ...
}
```

We go ahead and slide this AR forecaster over the working `epi_df` of COVID-19
cases. Note that we actually model the `cases_7dav` column, to operate on the
scale of smoothed COVID-19 cases. This is clearly equivalent, up to a constant,
to modeling weekly sums of COVID-19 cases.

```{r, warning=FALSE}
fc_time_values <- seq(
  from = as.Date("2020-06-01"),
  to = as.Date("2021-12-01"),
  by = "1 months")

fcasts <- epi_slide(
  x,
  .f = ~ arx_forecaster(
    epi_data = .x,
    outcome = "cases_7dav",
    predictors = "cases_7dav",
    trainer = quantile_reg(),
    args_list = arx_args_list(ahead = 7))$predictions,
  .window_size = 120,
  .ref_time_values = fc_time_values)

# grab just the relevant columns, and make them easier to plot
fcasts <- fcasts %>%
  select(geo_value, time_value, cases_7dav, .pred, .pred_distn) %>%
  pivot_quantiles_wider(".pred_distn")
fcasts
```

Note that we have used the argument `.ref_time_values` to compute the forecast
at a specific subset of reference time values. We get out 4 new columns:
`fc_target_date`, `0.05`, `0.5`, `0.95` that correspond to the date the forecast
is for (rather than the date it was made on), the point forecast, and the lower
and upper endpoints of the 95\% prediction band.[^1]

[^1]: If instead we had set `as_list_col = TRUE` in the call to `epi_slide()`,
then we would have gotten a list column `fc`, where each element of `fc`
contains these results.

To finish off, we plot the forecasts at some times (spaced out by a few months)
over the last year, at multiple horizons: 7, 14, 21, and 28 days ahead. To do
so, we encapsulate the process of generating forecasts into a simple function,
so that we can call it a few times.

```{r, message = FALSE, warning = FALSE}
k_week_ahead <- function(ahead = 7) {
  epi_slide(
    x,
    ~ arx_forecaster(
      epi_data = .x,
      outcome = "cases_7dav",
      predictors = "cases_7dav",
      trainer = quantile_reg(),
      args_list = arx_args_list(ahead = ahead))$predictions,
    .window_size = 120,
    .ref_time_values = fc_time_values) %>%
    select(geo_value, time_value, cases_7dav, .pred, .pred_distn) %>%
    pivot_quantiles_wider(".pred_distn")
}

# First generate the forecasts, and bind them together
z <- map(c(7, 14, 21, 28), k_week_ahead) %>% list_rbind()
```

```{r, fig.width=8, fig.height=9}
#| code-fold: true
ggplot(z) +
  geom_line(data = x, aes(x = time_value, y = cases_7dav), color = "gray50") +
  geom_ribbon(aes(x = time_value, ymin = `0.05`, ymax = `0.95`,
                  group = time_value, fill = geo_value), alpha = 0.4) +
  geom_line(aes(x = time_value, y = `0.5`, group = time_value)) +
  geom_point(aes(x = time_value, y = `0.5`, group = time_value), size = 0.5) +
  #geom_vline(data = tibble(x = fc_time_values), aes(xintercept = x),
  #           linetype = 2, alpha = 0.5) +
  facet_wrap(vars(geo_value), scales = "free_y", nrow = 3) +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  scale_x_date(minor_breaks = "1 months", date_labels = "%b %Y") +
  scale_fill_viridis_d(guide = "none", end = .9) +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

Two points are worth making. First, the AR model's performance here is pretty
spotty. At various points in time, we can see that its forecasts are volatile
(its point predictions are all over the place), or overconfident (its bands are
too narrow), or both at the same time. This is only meant as a simple demo and
not entirely unexpected given the way the AR model is set up. The
[`epipredict`](https://cmu-delphi.github.io/epipredict) package,
offers a suite of predictive modeling tools
that improve on many of the shortcomings of the above simple AR model (simply
using all states for training rather than 6 is a huge improvement).

Second, the AR forecaster here is using finalized data, meaning, it uses the
latest versions of signal values (reported COVID-19 cases) available, for both
training models and making predictions historically. However, this is not
reflective of the provisional nature of the data that it must cope with in a
true forecast task. Training and making predictions on finalized data can lead
to an overly optimistic sense of accuracy; see, for example,
[@McDonaldBien2021] and references
therein. Fortunately, the `epiprocess` package provides a data structure called
`epi_archive` that can be used to store all data revisions, and furthermore, an
`epi_archive` object knows how to slide computations in the correct
version-aware sense (for the computation at each reference time $t$, it uses
only data that would have been available as of $t$). We will revisit this
example in the [archive
vignette](https://cmu-delphi.github.io/epiprocess/articles/archive.html).
