# Sliding version-unaware and version-aware ARX forecasters across dates

```{r}
#| echo: false
source("_common.R")
```

A key function from the epiprocess package is `epix_slide()` (refer to the
following vignette for the basics of the function: ["Work with archive objects
and data
revisions"](https://cmu-delphi.github.io/epiprocess/articles/archive.html))
which allows performing version-aware computations. That is, the function only
uses data that would have been available as of time t for that reference time.

In this vignette, we use `epix_slide()` for backtesting our `arx_forecaster` on
historical COVID-19 case data from the US and from Canada. We first examine the
results from a version-unaware forecaster, comparing two different fitting
engines and then we contrast this with version-aware forecasting. The former
will proceed by constructing an `epi_archive` that erases its version
information and then use `epix_slide()` to forecast the future. The latter will
keep the versioned data and proceed similarly by using `epix_slide()` to
forecast the future.

## Version-unaware forecasting

### Example using CLI and case data from US states

First, we download the version history (i.e. archive) of the percentage of
doctor’s visits with CLI (COVID-like illness) computed from medical insurance
claims and the number of new confirmed COVID-19 cases per 100,000 population
(daily) for all 50 states from the COVIDcast API. We process as before, with the
modification that we use `sync = "locf"` in `epix_merge()` so that the last
version of each observation can be carried forward to extrapolate unavailable
versions for the less up-to-date input archive.

```{r grab-epi-data}
us_raw_history_dfs <- readRDS(url(
  "https://github.com/cmu-delphi/epipredict/raw/dev/vignettes/articles/all_states_covidcast_signals.rds"
))

us_cli_archive <- us_raw_history_dfs[[1]] %>%
  select(geo_value, time_value, version = issue, percent_cli = value) %>%
  as_epi_archive(compactify = TRUE)
us_cases_archive <- us_raw_history_dfs[[2]] %>%
  select(geo_value, time_value, version = issue, case_rate = value) %>%
  as_epi_archive(compactify = TRUE)

us_archive <- epix_merge(
  us_cli_archive, us_cases_archive,
  sync = "locf", compactify = TRUE
)
```

We then get latest snapshot of the data from the archive by using
`epix_as_of()`. We then create fake version information by setting `version =
time_value`. This creates an archive that pretends to have the latest data
available (since at version time `x` it has all the data up to time_value `x`,
which in reality is unrealistic because the time values of the data received at
version time `x` often lags by a few days, not to mention the later corrections
that are amended to the data).

```{r make-arx-kweek, warning = FALSE}
# Get latest snapshot of data and pretend it's an archive
us_latest <- us_archive %>%
  epix_as_of(version = max(.$versions_end)) %>%
  mutate(version = time_value) %>%
  as_epi_archive()
fc_time_values <- seq(
  from = as.Date("2020-08-01"),
  to = as.Date("2021-11-01"),
  by = "1 month"
)
aheads <- c(7, 14, 21, 28)

forecast_k_week_ahead <- function(epi_archive, outcome, predictors, ahead = 7, engine) {
  epi_archive %>%
    epix_slide(
      .f = function(x, gk, rtv) {
        arx_forecaster(
          x, outcome, predictors, engine,
          args_list = arx_args_list(ahead = ahead)
        )$predictions
      },
      .before = 120,
      .versions = fc_time_values
    ) %>%
      mutate(engine_type = engine$engine) %>%
      pivot_quantiles_wider(.pred_distn)
}

# Generate the forecasts and bind them together
forecasts_version_unaware <- bind_rows(
  map(aheads, ~ forecast_k_week_ahead(
    us_latest,
    outcome = "case_rate",
    predictors = c("case_rate", "percent_cli"),
    ahead = .x,
    engine = linear_reg()
  )),
  map(aheads, ~ forecast_k_week_ahead(
    us_latest,
    outcome = "case_rate",
    predictors = c("case_rate", "percent_cli"),
    ahead = .x,
    engine = rand_forest(mode = "regression")
  ))
)
```

Here, `arx_forecaster()` does all the heavy lifting. It creates leads of the
target (respecting time stamps and locations) along with lags of the features
(here, the response and doctors visits), estimates a forecasting model using the
specified engine, creates predictions, and non-parametric confidence bands.

To see how the predictions compare, we plot them on top of the latest case
rates. Note that even though we've fitted the model on all states, we'll just
display the results for two states, California (CA) and Florida (FL), to get a
sense of the model performance while keeping the graphic simple.

```{r plot-arx, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
#| code-fold: true
forecasts_filtered <- forecasts_version_unaware %>%
  tibble() %>%
  filter(geo_value %in% c("ca", "fl"))
latest_data_filtered <- us_latest$DT %>%
  tibble() %>%
  filter(geo_value %in% c("ca", "fl"))

ggplot(forecasts_filtered, aes(x = target_date, group = forecast_date, fill = engine_type)) +
  geom_line(
    data = latest_data_filtered, aes(x = time_value, y = case_rate),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`), alpha = 0.4) +
  geom_line(aes(y = .pred)) +
  geom_point(aes(y = .pred), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_grid(engine_type ~ geo_value, scales = "free") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  labs(x = "Date", y = "Reported COVID-19 case rates") +
  theme(legend.position = "none")
```

For the two states of interest, simple linear regression clearly performs better
than random forest in terms of accuracy of the predictions and does not result
in such in overconfident predictions (overly narrow confidence bands). Though,
in general, neither approach produces amazingly accurate forecasts. This could
be because the behaviour is rather different across states and the effects of
other notable factors such as age and public health measures may be important to
account for in such forecasting. Including such factors as well as making
enhancements such as correcting for outliers are some improvements one could
make to this simple model.[^1]

[^1]: Note that, despite the above caveats, simple models like this tend to
out-perform many far more complicated models in the online Covid forecasting due
to those models high variance predictions.

### Example using case data from Canada

By leveraging the flexibility of `epiprocess`, we can apply the same techniques
to data from other sources. Since some collaborators are in British Columbia,
Canada, we'll do essentially the same thing for Canada as we did above.

The [COVID-19 Canada Open Data Working Group](https://opencovid.ca/) collects
daily time series data on COVID-19 cases, deaths, recoveries, testing and
vaccinations at the health region and province levels. Data are collected from
publicly available sources such as government datasets and news releases.
Unfortunately, there is no simple versioned source, so we have created our own
from the Github commit history.

First, we load versioned case rates at the provincial level. After converting
these to 7-day averages (due to highly variable provincial reporting
mismatches), we then convert the data to an `epi_archive` object, and extract
the latest version from it. Finally, we run the same forcasting exercise as for
the American data, but here we compare the forecasts produced from using simple
linear regression with those from using boosted regression trees.

```{r get-can-fc, warning = FALSE}
# source("drafts/canada-case-rates.R)
can <- epidatasets::can_prov_cases
can <- can %>%
  group_by(version, geo_value) %>%
  arrange(time_value) %>%
  mutate(cr_7dav = RcppRoll::roll_meanr(case_rate, n = 7L)) %>%
  as_epi_archive(compactify = TRUE)

can_latest <- epix_as_of(can, max_version = max(can$DT$version)) %>%
  mutate(version = time_value) %>%
  as_epi_archive()

# Generate the forecasts, and bind them together
can_fc <- bind_rows(
  map(aheads,
    ~ forecast_k_week_ahead(can_latest, "cr_7dav", "cr_7dav", .x, linear_reg())
  ),
  map(aheads,
    ~ forecast_k_week_ahead(can_latest, "cr_7dav", "cr_7dav", .x, boost_tree(mode = "regression", trees = 20))
  )
)
```

The first figure shows the results for all of the provinces using linear regression.

```{r plot-can-fc-lr, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 12}
#| code-fold: true
ggplot(
  can_fc %>% filter(engine_type == "lm"),
  aes(x = target_date, group = forecast_date)
) +
  coord_cartesian(xlim = lubridate::ymd(c("2020-12-01", NA))) +
  geom_line(
    data = can_latest$DT %>% tibble(), aes(x = time_value, y = cr_7dav),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = geo_value),
    alpha = 0.4
  ) +
  geom_line(aes(y = .pred)) +
  geom_point(aes(y = .pred), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 3) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  labs(
    title = "Using simple linear regression", x = "Date",
    y = "Reported COVID-19 case rates"
  ) +
  theme(legend.position = "none")
```

Compare those forecasts with a related set using Gradient Boosting.

```{r plot-can-fc-boost, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 12}
#| code-fold: true
ggplot(
  can_fc %>% filter(engine_type == "xgboost"),
  aes(x = target_date, group = forecast_date)
) +
  coord_cartesian(xlim = lubridate::ymd(c("2020-12-01", NA))) +
  geom_line(
    data = can_latest$DT %>% tibble(), aes(x = time_value, y = cr_7dav),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = geo_value),
    alpha = 0.4
  ) +
  geom_line(aes(y = .pred)) +
  geom_point(aes(y = .pred), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 3) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  labs(
    title = "Using boosted regression trees", x = "Date",
    y = "Reported COVID-19 case rates"
  ) +
  theme(legend.position = "none")
```

Both approaches tend to produce quite volatile forecasts (point predictions)
and/or are overly confident (very narrow bands), particularly when boosted
regression trees are used. But as this is meant to be a simple demonstration of
sliding with different engines in `arx_forecaster`, we may devote another
vignette to work on improving the predictive modelling using the suite of tools
available in epipredict.

## Version-aware forecasting

### Example using case data from US states

We will now run pseudoprospective forecasts based on properly-versioned data
(that would have been available in real-time) to forecast future COVID-19 case
rates from current and past COVID-19 case rates for all states. All we have to
do is use the historical archive of the data with version information,
`us_archive`, instead of `us_latest` like we did above, in the argument to our
forecaster wrapper `forecast_k_week_ahead()`. Below we do that computation, tag
it, and combine it with the forecasts from one of the engines made above.

```{r make-faithful-forecast}
forecasts_version_aware <- map(aheads, ~ forecast_k_week_ahead(
    us_archive,
    outcome = "case_rate",
    predictors = c("case_rate", "percent_cli"),
    ahead = .x,
    engine = linear_reg()
  )) %>%
    bind_rows() %>%
    mutate(version = "version faithful")
```

Now we can plot the results on top of the latest case rates. As before, we will
only display and focus on the results for FL and CA for simplicity.

```{r plot-ar-asof, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
#| code-fold: true
forecasts_filtered <- bind_rows(
    forecasts_version_aware,
    forecasts_version_unaware %>%
      filter(engine_type == "lm") %>%
      mutate(version = "version unfaithful")
  ) %>%
  tibble() %>%
  filter(geo_value %in% c("ca", "fl"))
latest_data_filtered <- us_latest$DT %>%
  tibble() %>%
  select(-version) %>%
  filter(geo_value %in% c("ca", "fl"))

ggplot(forecasts_filtered, aes(x = target_date, group = forecast_date, fill = version)) +
  geom_line(
    data = latest_data_filtered, aes(x = time_value, y = case_rate),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`), alpha = 0.4) +
  geom_line(aes(y = .pred)) +
  geom_point(aes(y = .pred), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_grid(version ~ geo_value, scales = "free") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  labs(x = "Date", y = "Reported COVID-19 case rates") +
  theme(legend.position = "none")
```

Again, we observe that the results are not great for these two states, but
that's likely due to the simplicity of the model (ex. the omission of key
factors such as age and public health measures) and the quality of the data (ex.
we have not personally corrected for anomalies in the data).

We shall leave it to the reader to try the above version aware and unaware
forecasting exercise on the Canadian case rate data. The above code for the
American state data should be readily adaptable for this purpose.
